---
title: 'P&S-2022: Lab assignment 2'
author: "Name1, Name2, Name3"
output:
  html_document:
    df_print: paged
---

## General comments and instructions

-   Complete solution will give you **4 points** (working code with explanations + oral defense). Submission deadline **November 1, 2023, 22:00**\
-   The report must be prepared as an *R notebook*; you must submit to **cms** both the source *R notebook* **and** the generated html file\
-   At the beginning of the notebook, provide a work-breakdown structure estimating efforts of each team member\
-   For each task, include
    -   problem formulation and discussion (what is a reasonable answer to discuss);\
    -   the corresponding $\mathbf{R}$ code with comments (usually it is just a couple of lines long);\
    -   the statistics obtained (like sample mean or anything else you use to complete the task) as well as histograms etc to illustrate your findings;\
    -   justification of your solution (e.g. refer to the corresponding theorems from probability theory);\
    -   conclusions (e.g. how reliable your answer is, does it agree with common sense expectations etc)\
-   The **team id number** referred to in tasks is the **two-digit** ordinal number of your team on the list. Include the line **set.seed(team id number)** at the beginning of your code to make your calculations reproducible. Also observe that the answers **do** depend on this number!\
-   Take into account that not complying with these instructions may result in point deduction regardless of whether or not your implementation is correct.

------------------------------------------------------------------------

### Task 1

#### In this task, we discuss the $[7,4]$ Hamming code and investigate its reliability. That coding system can correct single errors in the transmission of $4$-bit messages and proceeds as follows:

-   given a message $\mathbf{m} = (a_1 a_2 a_3 a_4)$, we first encode it to a $7$-bit *codeword* $\mathbf{c} = \mathbf{m}G = (x_1 x_2 x_3 x_4 x_5 x_6 x_7)$, where $G$ is a $4\times 7$ *generator* matrix

-   the codeword $\mathbf{c}$ is transmitted, and $\mathbf{r}$ is the received message $\mathbf{r}$ is checked for errors by calculating the *syndrome vector* $\mathbf{z} := \mathbf{r} H$, for a $7 \times 3$ *parity-check* matrix $H$\

-   if a single error has occurred in $\mathbf{r}$, then the binary $\mathbf{z} = (z_1 z_2 z_3)$ identifies the wrong bit no. $z_1 + 2 z_2 + 4z_3$; thus $(0 0 0)$ shows there was no error (or more than one), while $(1 1 0 )$ means the third bit (or more than one) got corrupted\

-   if the error was identified, then we flip the corresponding bit in $\mathbf{r}$ to get the corrected $\mathbf{r}^* = (r_1 r_2 r_3 r_4 r_5 r_6 r_7)$;\

-   the decoded message is then $\mathbf{m}^*:= (r_3r_5r_6r_7)$.

#### The **generator** matrix $G$ and the **parity-check** matrix $H$ are given by

$$  
    G := 
    \begin{pmatrix}
        1 & 1 & 1 & 0 & 0 & 0 & 0 \\
        1 & 0 & 0 & 1 & 1 & 0 & 0 \\
        0 & 1 & 0 & 1 & 0 & 1 & 0 \\
        1 & 1 & 0 & 1 & 0 & 0 & 1 \\
    \end{pmatrix},
 \qquad 
    H^\top := \begin{pmatrix}
        1 & 0 & 1 & 0 & 1 & 0 & 1 \\
        0 & 1 & 1 & 0 & 0 & 1 & 1 \\
        0 & 0 & 0 & 1 & 1 & 1 & 1
    \end{pmatrix}
$$

#### Assume that each bit in the transmission $\mathbf{c} \mapsto \mathbf{r}$ gets corrupted independently of the others with probability $p = \mathtt{id}/100$, where $\mathtt{id}$ is your team number. Your task is the following one.

1.  Simulate the encoding-transmission-decoding process $N$ times and find the estimate $\hat p$ of the probability $p^*$ of correct transmission of a single message $\mathbf{m}$. Comment why, for large $N$, $\hat p$ is expected to be close to $p^*$.\
2.  By estimating the standard deviation of the corresponding indicator of success by the standard error of your sample and using the CLT, predict the \emph{confidence} interval $(p^*-\varepsilon, p^* + \varepsilon)$, in which the estimate $\hat p$ falls with probability at least $0.95$.\
3.  What choice of $N$ guarantees that $\varepsilon \le 0.03$?\
4.  Draw the histogram of the number $k = 0,1,2,3,4$ of errors while transmitting a $4$-digit binary message. Do you think it is one of the known distributions?

#### You can (but do not have to) use the chunks we prepared for you

#### First, we set the **id** of the team and define the probability $p$ and the generator and parity-check matrices $G$ and $H$

```{r}
# your team id number 
                          ###
id <- 5                 ### Change to the correct id!
                          ###
set.seed(id)
p <- id/100
# matrices G and H
G <- matrix(c(1, 1, 1, 0, 0, 0, 0,
		1, 0, 0, 1, 1, 0, 0,
		0, 1, 0, 1, 0, 1, 0,
		1, 1, 0, 1, 0, 0, 1), nrow = 4, byrow = TRUE)
H <- t(matrix(c(1, 0, 1, 0, 1, 0, 1,
		0, 1, 1, 0, 0, 1, 1,
		0, 0, 0, 1, 1, 1, 1), nrow = 3, byrow = TRUE))
cat("The matrix G is: \n")
G
cat("The matrix H is: \n")
H
cat("The product GH must be zero: \n")
(G%*%H) %%2
```

```{r}
encode <- function(m) {
  return(m %*% G %% 2)
}

apply_errors <- function(m, e) {
  return((m + e) %% 2)
}

decode <- function(m) {
  decode_row <- function(row) {
    s <- (row %*% H) %% 2
    n <- 3
    dm <- sum(s * (2^(0:(n-1))))
    if (dm != 0) {
      row[dm] <- 1 - row[dm]  # flip bit
    }
    return(row[c(3, 5, 6, 7)])
  }
    return(t(apply(m, 1, decode_row)))
}

m <- matrix(c(0, 1, 0, 1), nrow = 1)
e <- matrix(c(0, 0, 0, 1, 0, 0, 0), nrow = 1)
c <- encode(m)
ec <- apply_errors(c, e)
d <- decode(ec)
cat("Initial message: ", m, "\n")
cat("Encoded message: ", c, "\n")
cat("Encoded message after errors were applied: ", ec, "\n")
cat("Decoded message: ", d, "\n")
```

```{r}
message_generator <- function(N) {
  matrix(sample(c(0,1), 4*N, replace = TRUE), nrow = N)
}

error_generator <- function(N) {
  matrix(sample(c(0,1), 7*N, replace = TRUE, prob = c(1-p, p)), nrow = N)
}
```

The next steps include detecting the errors in the received messages, correcting them, and then decoding the obtained messages. After this, you can continue with calculating all the quantities of interest

1.  Simulate the encoding-transmission-decoding process N times and find the estimate $\hat{p}$ of the probability $p∗$ of correct transmission of a single message m. Comment why, for large N , $\hat{p}$ is expected to be close to $p∗$

```{r}
N <- 5000000

messages <- message_generator(N) # Generate N messages
codewords <- encode(messages) # Encode messages
errors <- error_generator(N) # Generate errors
received <- apply_errors(codewords, errors) # Apply errors to get received messages with errors
decoded_messages <- decode(received) # Decode messages
```

```{r}
incorrectly_decoded_messages <- sum(apply(xor(messages, decoded_messages), 1, any))
p_hat = incorrectly_decoded_messages/N
print(p_hat)
```

Let each transmission of a message be a Bernoulli trial, where $$
X_i = 
\begin{cases} 
1 & \text{if the $i$-th message is decoded correctly,} \\
0 & \text{otherwise.}
\end{cases}
$$ All random variables $X_i$ is independent and identically distributed.

Let $(x_1, x_2, ... , x_N)$ be realization of $(X_1, X_2, ... , X_N)$, then $\hat{p}$ is estimated as $\hat{p}=\frac{x_1 + ... + x_N}{N}$ for $N$ independent transitions.

Then the true probability of the correct decoding is $p* = E(X_i)$

Message can be decoded incorrectly only if there were 1 or 0 errors in it. Each encoded message consist of 7 bits, so probability of 2 or more errors in it is $p* = E(X_i) = 1-(1-p)^7-\binom{7}{1}p(1-p)^6 \approx 0.04438$, $p$ is probability of an error.

By the strong law of large the sample mean of i.i.d. random variables converges almost surely to the expected value as $N \to \infty$

$$
\hat{p} = \frac{1}{N} \sum_{i=1}^{N} X_i \xrightarrow{a.s.} \mathbb{E}[X_i] = p^*.
$$

2.  By estimating the standard deviation of the corresponding indicator of success by the standard error of your sample and using the CLT, find the half-length $ε$ of the confidence interval $(\hat{p} − ε, \hat{p} + ε)$, which contains the true value $p∗$ with probability at least $0.95$. What choice of N guarantees that $ε \le 0.03$?

Let $X_1, \dots, X_N$ be independent Bernoulli indicators of success with $$
\mathbb{P}(X_i = 1) = 1-p^*, \quad \mathbb{P}(X_i = 0) = p^*.
$$

$$
\mu=E(x_i)=1-p^*\\
\sigma^2 = E(X_i^2)-E(X_i)^2=p^*(1-p^*)
$$ The sample mean is $$
\hat{p} = \frac{1}{N}\sum_{i=1}^{N} X_i.
$$

By the Central Limit Theorem, for large $N$, $$
\hat{p} \sim \mathcal{N}\Big(\mu, \sigma^2\Big) \Rightarrow \hat{p} \sim \mathcal{N}\Big(p^*, \frac{p^*(1-p^*)}{N}\Big)
$$ A $95\%$ confidence interval is $$
(\hat{p} - z_{0.975} \cdot \mathrm{SE}(\hat{p}); \; \hat{p} + z_{0.975} \cdot \mathrm{SE}(\hat{p}))
$$ $z_{0.975}$ is chosen because in the standard normal distribution 95% of area under curve is between $x=-z_{0.975}$ and $x=z_{0.975}$

where the standard error is estimated by $$
\mathrm{SE}(\hat{p}) \approx \sqrt{\frac{\hat{p}(1-\hat{p})}{N}}, \quad z_{0.975} \approx 1.96.
$$ From this $\varepsilon$ is equal to $$
\varepsilon = z_{0.975} \sqrt{\frac{\hat{p}(1-\hat{p})}{N}}.
$$

To guarantee $\varepsilon \le 0.03$, we need to solve solve for $N$: $$
N \ge \frac{\hat{p}(1-\hat{p})}{(\varepsilon / z_{0.975})^2}.
$$ To guarantee that $N$ is greater or equal we need to use maximum possible value for $\hat{p} (1-\hat{p})$ which is $0.25$ at point $0.5$ (can be found by finding derivative and solving equation where it is equal to 0). By plugging $\varepsilon=0.03$, $z_{0.975} \approx 1.96$ (from table) and $\hat{p}(1-\hat{p}) \le 0.25$, we get $$
N \ge \frac{0.25}{(0.03/1.96)^2} \approx 1067.
$$

3.  Draw the histogram of the number k = 0, 1, 2, 3, 4 of errors while transmitting a 4-digit binary message. Do you think the random variable that counts the number of wrong bits in a decoded message has one of the known distributions? Justify your answer.

```{r}
number_of_errors <- apply(xor(messages, decoded_messages), 1, sum)

# Normal barplot
barplot(table(number_of_errors),
        main = "Barplot of Bit Errors",
        xlab = "Number of Errors")

# Zoomed barplot
barplot(table(number_of_errors),
        main = "Barplot of Bit Errors (Zoomed)",
        xlab = "Number of Errors",
        ylim = c(0, 100000))

# Logarithmically scaled barplot
barplot(table(number_of_errors),
        main = "Barplot of Bit Errors (log scale)",
        xlab = "Number of Errors",
        log="y")

print(table(number_of_errors))
```

I don't think it is one of the known distributions because the number of errors in the decoded message depends on both the errors occurring during transmission and the logic of correction. Also, not only the number of flipped bits impacts the number of errors, but also the positions of the flipped bits. The number of errors during transmission follows a Binomial distribution with parameters $7$ and $p$ ($B(7,\,p)$). If more than one error occurs, during the correction step there are several cases: the syndrome vector may be zero (no bits flipped) or non-zero, in which case some bit is flipped (this could be a correct or incorrect bit). This logic cannot be represented as any known distribution. The combination of the Binomial distribution of transmission errors and the decoding logic therefore also cannot be represented as a standard distribution.

**Conclusion**

In this task, we simulated encoding, transmitting, and decoding of messages using Hamming(7, 4) codes, and also calculated the probability of errors. A plot showing the number of cases with different numbers of errors was produced. Simulation results were consistent with our expectations: when the probability of a bit flip during transmission is small (e.g., $0.05$), most messages are received without error or with a single error, which can be corrected. The probability of more errors is quite small (in our case $\approx 0.04438$), showing that Hamming encoding can be quite reliable (at least in some cases). One thing worth noting is that in the simulated system, each bit can be flipped at most once (for instance, a double error cannot flip the same bit twice and leave it unchanged), which may not be true for some real-world systems.
$$$$
------------------------------------------------------------------------

### Task 2.

In this task, we discuss a real-life process that is well modelled by a Poisson distribution. As you remember, a Poisson random variable describes occurrences of rare events, i.e., counts the number of successes in a large number of independent random experiments. One of the typical examples is the **radioactive decay** process.

Consider a sample of radioactive element of mass $m$, which has a big *half-life period* $T$; it is vitally important to know the probability that during a one second period, the number of nuclei decays will not exceed some critical level $k$. This probability can easily be estimated using the fact that, given the *activity* ${\lambda}$ of the element (i.e., the probability that exactly one nucleus decays in one second) and the number $N$ of atoms in the sample, the random number of decays within a second is well modelled by Poisson distribution with parameter $\mu:=N\lambda$. Next, for the sample of mass $m$, the number of atoms is $N = \frac{m}{M} N_A$, where $N_A = 6 \times 10^{23}$ is the Avogadro constant, and $M$ is the molar (atomic) mass of the element. The activity of the element, $\lambda$, is $\log(2)/T$, where $T$ is measured in seconds.

Assume that a medical laboratory receives $n$ samples of radioactive element ${{}^{137}}\mathtt{Cs}$ (used in radiotherapy) with half-life period $T = 30.1$ years and mass $m = \mathtt{team\, id \,number} \times 10^{-6}$ g each. Denote by $X_1,X_2,\dots,X_n$ the **i.i.d. r.v.**'s counting the number of decays in sample $i$ in one second.

1.  Specify the parameter of the Poisson distribution of $X_i$ (you'll need the atomic mass of *Cesium-137*)\
2.  Show that the distribution of the sample means of $X_1,\dots,X_n$ gets very close to a normal one as $n$ becomes large and identify that normal distribution. To this end,
    -   simulate the realization $x_1,x_2,\dots,x_n$ of the $X_i$ and calculate the sample mean $s=\overline{\mathbf{x}}$;
    -   repeat this $K$ times to get the sample $\mathbf{s}=(s_1,\dots,s_K)$ of means and form the empirical cumulative distribution function $\hat F_{\mathbf{s}}$ of $\mathbf{s}$;
    -   identify $\mu$ and $\sigma^2$ such that the \textbf{c.d.f.} $F$ of $\mathscr{N}(\mu,\sigma^2)$ is close to the \textbf{e.c.d.f.} $\hat F_{\mathbf{s}}$ and plot both **c.d.f.**'s on one graph to visualize their proximity (use the proper scales!);
    -   calculate the maximal difference between the two \textbf{c.d.f.}'s;
    -   consider cases $n = 5$, $n = 10$, $n=50$ and comment on the results.\
3.  Calculate the largest possible value of $n$, for which the total number of decays in one second is less than $8 \times 10^8$ with probability at least $0.95$. To this end,
    -   obtain the theoretical bound on $n$ using Markov inequality, Chernoff bound and Central Limit Theorem, and compare the results;\
    -   simulate the realization $x_1,x_2,\dots,x_n$ of the $X_i$ and calculate the sum $s=x_1 + \cdots +x_n$;
    -   repeat this $K$ times to get the sample $\mathbf{s}=(s_1,\dots,s_K)$ of sums;
    -   calculate the number of elements of the sample which are less than critical value ($8 \times 10^8$) and calculate the empirical probability; comment whether it is close to the desired level $0.95$

Task1:
```{r}
# Константи
m <- 5e-6              # маса зразка в грамах
M <- 137               # молярна маса Cs-137 в г/моль
N_A <- 6e23            # число Авогадро
T_years <- 30.1        # період напіврозпаду в роках

# Переводимо період напіврозпаду в секунди
T <- T_years * 365.25 * 24 * 3600

# Кількість атомів у зразку
N <- (m / M) * N_A

# Активність (ймовірність розпаду одного атома за секунду)
lambda <- log(2) / T

# Параметр розподілу Пуассона
mu <- N * lambda

# Виводимо результати
cat("Кількість атомів N =", format(N, scientific = TRUE), "\n")
cat("Активність λ =", format(lambda, scientific = TRUE), "1/с\n")
cat("Параметр Пуассона μ =", format(mu, scientific = TRUE), "\n")
```
Task 2:
```{r}
# Використовуємо μ з попереднього кроку
mu <- 15979224
# Функція для симуляції та покращеного аналізу
simulate_and_analyze_improved <- function(n, mu, K) {
  # Симуляція K вибіркових середніх
  sample_means <- replicate(K, {
    x <- rpois(n, lambda = mu)
    mean(x)
  })
  
  # Теоретичні параметри нормального розподілу
  mu_theory <- mu
  sigma_theory <- sqrt(mu / n)
  
  # Емпірична функція розподілу
  Fs <- ecdf(sample_means)
  
  # ======== ГРАФІК 1: ГІСТОГРАМА + НОРМАЛЬНА ЩІЛЬНІСТЬ ========
  # Діапазон для графіків
  xlims <- c(mu_theory - 4*sigma_theory, mu_theory + 4*sigma_theory)
  
  # Гістограма з щільністю
  hist(sample_means, 
       probability = TRUE,  # щільність, а не частоти
       breaks = 60,
       col = rgb(0, 0, 1, 0.3),  # напівпрозорий синій
       border = "blue",
       xlim = xlims,
       main = paste("Розподіл вибіркових середніх для n =", n),
       xlab = "Вибіркове середнє",
       ylab = "Щільність")
  
  # Накладаємо теоретичну нормальну щільність
  curve(dnorm(x, mean = mu_theory, sd = sigma_theory),
        col = "red",
        lwd = 3,
        add = TRUE)
  
  legend("topright",
         legend = c("Емпірична щільність", "Теоретична N(μ, σ²)"),
         fill = c(rgb(0, 0, 1, 0.3), NA),
         border = c("blue", NA),
         lty = c(NA, 1),
         col = c(NA, "red"),
         lwd = c(NA, 3))
  
  # ======== ГРАФІК 2: CDF З ПРАВИЛЬНИМ МАСШТАБОМ ========
  # Знаходимо область з найбільшими змінами (μ ± 3σ)
  xlims_cdf <- c(mu_theory - 3*sigma_theory, mu_theory + 3*sigma_theory)
  
  plot(Fs,
       xlim = xlims_cdf,
       ylim = c(0, 1),
       col = "blue",
       lwd = 2,
       main = paste("Порівняння CDF для n =", n),
       xlab = "Вибіркове середнє",
       ylab = "F(x)",
       cex.main = 1.2)
  
  # Теоретична CDF
  curve(pnorm(x, mean = mu_theory, sd = sigma_theory),
        col = "red",
        lwd = 2,
        lty = 2,  # пунктирна лінія для кращого розрізнення
        add = TRUE)
  
  legend("topleft",
         legend = c("Емпірична F̂ₛ", "Теоретична F"),
         col = c("blue", "red"),
         lwd = 2,
         lty = c(1, 2))

  # Обчислюємо максимальну різницю
  x_sorted <- sort(sample_means)
  empirical_cdf <- Fs(x_sorted)
  theoretical_cdf <- pnorm(x_sorted, mean = mu_theory, sd = sigma_theory)
  max_diff <- max(abs(empirical_cdf - theoretical_cdf))

  # Виводимо результати
  cat("\n========================================\n")
  cat("Результати для n =", n, "\n")
  cat("========================================\n")
  cat("Теоретичне μ =", mu_theory, "\n")
  cat("Теоретичне σ =", sigma_theory, "\n")
  cat("Емпіричне середнє =", mean(sample_means), "\n")
  cat("Емпіричне σ =", sd(sample_means), "\n")
  cat("Максимальна різниця |F̂ₛ - F| =", max_diff, "\n")
  cat("========================================\n")
  
  return(list(
    sample_means = sample_means,
    mu_theory = mu_theory,
    sigma_theory = sigma_theory,
    max_diff = max_diff
  ))
}

# ======== ВИКОНАННЯ ========
set.seed(52)
K <- 10000
n_values <- c(5, 10, 50)

# Для кожного n створюємо 3 графіки
for (n in n_values) {
  # Встановлюємо layout для 3 графіків в один рядок
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
  
  cat("\n\n")
  cat(rep("=", 60), "\n")
  cat("АНАЛІЗ ДЛЯ n =", n, "\n")
  cat(rep("=", 60), "\n")
  result <- simulate_and_analyze_improved(n, mu, K)
  
  # Повертаємо стандартний layout
  par(mfrow = c(1, 1))
}
```
**Коментар про результати**:
- При n = 5: різниця ≈ 0.013, наближення вже досить добре
- При n = 10: різниця зменшується - 0.01, наближення покращується
- При n = 50: різниця мінімальна - 0.0058, наближення відмінне
- Висновок: Зі збільшенням n розподіл вибіркового середнього все краще наближається до нормального, що підтверджує ЦГТ

Task3:

```{r}
# Константи
critical_value <- 8e8  # 8 × 10^8
prob_threshold <- 0.05  # 1 - 0.95

# Підхід 1: Нерівність Маркова
# P(S > c) ≤ E[S]/c = nμ/c
# nμ/c ≤ 0.05  =>  n ≤ 0.05 × c / μ

n_markov <- floor((1 - prob_threshold) * critical_value / mu)

cat("============================================================\n")
cat("ПІДХІД 1: Нерівність Маркова\n")
cat("============================================================\n")
cat("Максимальне n (Марков) =", n_markov, "\n")
cat("E[S] = n × μ =", n_markov * mu, "\n")
cat("Теоретична верхня межа P(S > c) ≤", n_markov * mu / critical_value, "\n")
cat("\n")
```

```{r}
# Підхід 2: Нерівність Чернова
# Для S ~ Poisson(nμ), якщо c > nμ:
# P(S > c) ≤ exp(-nμ) × (e×nμ/c)^c  (точна формула складна)

# Використаємо спрощену: знаходимо n, при якому P(S > c) ≤ 0.05
# Шукаємо числово

chernoff_bound <- function(n, mu, c) {
  lambda <- n * mu
  if (c <= lambda) return(1)  # bound не працює
  
  # Chernoff bound для Poisson
  delta <- (c / lambda) - 1
  bound <- exp(-delta^2 * lambda / (2 + delta))
  return(bound)
}

# Бінарний пошук для знаходження максимального n
find_max_n_chernoff <- function(mu, c, threshold) {
  n_low <- 1
  n_high <- floor(c / mu) * 2  # верхня межа для пошуку
  
  while (n_high - n_low > 1) {
    n_mid <- floor((n_low + n_high) / 2)
    bound <- chernoff_bound(n_mid, mu, c)
    
    if (bound <= threshold) {
      n_low <- n_mid
    } else {
      n_high <- n_mid
    }
  }
  
  return(n_low)
}

n_chernoff <- find_max_n_chernoff(mu, critical_value, prob_threshold)

cat("============================================================\n")
cat("ПІДХІД 2: Нерівність Чернова\n")
cat("============================================================\n")
cat("Максимальне n (Чернов) =", n_chernoff, "\n")
cat("E[S] = n × μ =", n_chernoff * mu, "\n")
cat("Chernoff bound P(S > c) ≤", chernoff_bound(n_chernoff, mu, critical_value), "\n")
cat("\n")
```

```{r}
# Підхід 3: Центральна Гранична Теорема (нормальне наближення)
# S ~ N(nμ, nμ) приблизно
# P(S ≤ c) = Φ((c - nμ)/√(nμ)) ≥ 0.95
# (c - nμ)/√(nμ) ≥ Φ^(-1)(0.95) = 1.645

z_score <- qnorm(0.95)  # 1.644854

# Розв'язуємо квадратне рівняння:
# c - nμ ≥ z√(nμ)
# c - nμ ≥ z√μ × √n
# Нехай √n = t, тоді: nμ + z√μ × t - c ≤ 0
# μt² + z√μ × t - c ≤ 0

# Використаємо числовий пошук
clt_condition <- function(n, mu, c, z) {
  mean_s <- n * mu
  sd_s <- sqrt(n * mu)
  # P(S ≤ c) за нормальним наближенням
  prob <- pnorm(c, mean = mean_s, sd = sd_s)
  return(prob)
}

find_max_n_clt <- function(mu, c, target_prob) {
  n_low <- 1
  n_high <- floor(c / mu) * 2
  
  while (n_high - n_low > 1) {
    n_mid <- floor((n_low + n_high) / 2)
    prob <- clt_condition(n_mid, mu, c, z_score)
    
    if (prob >= target_prob) {
      n_low <- n_mid
    } else {
      n_high <- n_mid
    }
  }
  
  return(n_low)
}

n_clt <- find_max_n_clt(mu, critical_value, 0.95)

cat("============================================================\n")
cat("ПІДХІД 3: Центральна Гранична Теорема (ЦГТ)\n")
cat("============================================================\n")
cat("Максимальне n (ЦГТ) =", n_clt, "\n")
cat("E[S] = n × μ =", n_clt * mu, "\n")
cat("Теоретична P(S ≤ c) ≈", clt_condition(n_clt, mu, critical_value, z_score), "\n")
cat("\n")
```

```{r}
cat("============================================================\n")
cat("ПОРІВНЯННЯ ТЕОРЕТИЧНИХ РЕЗУЛЬТАТІВ\n")
cat("============================================================\n")
cat("Нерівність Маркова:  n ≤", n_markov, "\n")
cat("Нерівність Чернова:  n ≤", n_chernoff, "\n")
cat("ЦГТ (найточніше):    n ≤", n_clt, "\n")
cat("============================================================\n\n")
```

```{r}
# ======== СИМУЛЯЦІЯ ДЛЯ ПЕРЕВІРКИ ========

simulate_probability <- function(n, mu, critical_value, K = 10000) {
  # Лічильник випадків, коли сума ≤ critical_value
  success_count <- 0
  
  for (i in 1:K) {
    # Генеруємо n зразків з Poisson(μ)
    x <- rpois(n, lambda = mu)
    # Рахуємо суму
    s <- sum(x)
    # Перевіряємо умову
    if (s <= critical_value) {
      success_count <- success_count + 1
    }
  }
  
  # Емпірична ймовірність
  empirical_prob <- success_count / K
  
  return(list(
    n = n,
    empirical_prob = empirical_prob,
    success_count = success_count,
    total_simulations = K
  ))
}

# Налаштування
set.seed(52)
K_simulation <- 10000  # кількість симуляцій

cat("============================================================\n")
cat("СИМУЛЯЦІЙНА ПЕРЕВІРКА\n")
cat("============================================================\n")
cat("Кількість симуляцій K =", K_simulation, "\n")
cat("Критичне значення c = 8 × 10^8\n")
cat("Цільова ймовірність = 0.95\n\n")

# Перевіряємо всі три значення n
n_values_to_check <- c(n_markov, n_chernoff, n_clt)
n_values_to_check <- unique(n_values_to_check)  # прибираємо дублікати

results_simulation <- list()

for (n_test in n_values_to_check) {
  cat("------------------------------------------------------------\n")
  cat("Тестуємо n =", n_test, "\n")
  cat("------------------------------------------------------------\n")
  
  result <- simulate_probability(n_test, mu, critical_value, K_simulation)
  results_simulation[[as.character(n_test)]] <- result
  
  cat("Емпірична P(S ≤ c) =", result$empirical_prob, "\n")
  cat("Успішних випадків:", result$success_count, "з", result$total_simulations, "\n")
  
  # Порівняння з цільовою ймовірністю
  if (result$empirical_prob >= 0.95) {
    cat("✓ Умова ВИКОНАНА (≥ 0.95)\n")
  } else {
    cat("✗ Умова НЕ виконана (< 0.95)\n")
  }
  cat("\n")
}
```

```{r}
# ======== ГРАФІК: P(S ≤ c) від n ========

# Діапазон n навколо знайдених значень
n_range <- seq(max(1, n_clt - 5), n_clt + 5, by = 1)

# Теоретичні ймовірності (за ЦГТ)
theoretical_probs <- sapply(n_range, function(n) {
  mean_s <- n * mu
  sd_s <- sqrt(n * mu)
  pnorm(critical_value, mean = mean_s, sd = sd_s)
})

# Емпіричні ймовірності (симуляція для кількох точок)
n_sample <- seq(max(1, n_clt - 3), n_clt + 3, by = 1)
empirical_probs <- sapply(n_sample, function(n) {
  result <- simulate_probability(n, mu, critical_value, K = 5000)
  result$empirical_prob
})

# Графік
plot(n_range, theoretical_probs,
     type = "l",
     col = "red",
     lwd = 2,
     xlab = "n (кількість зразків)",
     ylab = "P(S ≤ 8×10⁸)",
     main = "Залежність ймовірності від n",
     ylim = c(0.85, 1))

# Додаємо емпіричні точки
points(n_sample, empirical_probs,
       col = "blue",
       pch = 19,
       cex = 1.5)

# Горизонтальна лінія для цільової ймовірності
abline(h = 0.95, col = "darkgreen", lty = 2, lwd = 2)

# Вертикальні лінії для знайдених n
abline(v = n_markov, col = "orange", lty = 3)
abline(v = n_clt, col = "purple", lty = 3)

legend("topright",
       legend = c("Теоретична (ЦГТ)", 
                  "Емпірична (симуляція)",
                  "Цільова P = 0.95",
                  paste("n (Марков) =", n_markov),
                  paste("n (ЦГТ) =", n_clt)),
       col = c("red", "blue", "darkgreen", "orange", "purple"),
       lty = c(1, NA, 2, 3, 3),
       pch = c(NA, 19, NA, NA, NA),
       lwd = c(2, NA, 2, 1, 1))

grid()
```

```{r}
cat("============================================================\n")
cat("ФІНАЛЬНЕ ПОРІВНЯННЯ\n")
cat("============================================================\n")
cat("Метод                  | Теоретичне n | Емпірична P(S≤c)\n")
cat("------------------------------------------------------------\n")
cat(sprintf("Нерівність Маркова     | n ≤ %2d       | %.4f\n", 
            n_markov, 
            results_simulation[[as.character(n_markov)]]$empirical_prob))
cat(sprintf("Нерівність Чернова     | n ≤ %2d       | %.4f\n", 
            n_chernoff, 
            results_simulation[[as.character(n_chernoff)]]$empirical_prob))
cat(sprintf("ЦГТ (найточніше)       | n ≤ %2d       | %.4f\n", 
            n_clt, 
            results_simulation[[as.character(n_clt)]]$empirical_prob))
cat("============================================================\n")
cat("\nВИСНОВОК: Максимальне n =", n_clt, "\n")
```
```{r}
# Перевірка:
n <- 50
expected_sum <- n * mu
cat("Очікувана сума E[S] =", expected_sum, "\n")
cat("Критичне значення =", critical_value, "\n")
cat("Різниця =", critical_value - expected_sum, "\n")
cat("Стандартне відхилення SD[S] =", sqrt(n * mu), "\n")
```


============================================================
ДЕТАЛЬНИЙ КОМЕНТАР ДО РЕЗУЛЬТАТІВ ПУНКТУ 3
============================================================

ТЕОРЕТИЧНІ РЕЗУЛЬТАТИ:
----------------------
Нерівність Маркова:  n ≤ 47
Нерівність Чернова:  n ≤ 50
ЦГТ (найточніше):    n ≤ 50

АНАЛІЗ ТА ПОЯСНЕННЯ РЕЗУЛЬТАТІВ:
---------------------------------

1. ЧОМУ ВСІ ТРИ МЕТОДИ ДАЛИ БЛИЗЬКІ РЕЗУЛЬТАТИ?

Параметр μ ≈ 1.6×10⁷ є дуже великим числом. Це означає, що в кожному 
окремому зразку відбувається приблизно 16 мільйонів радіоактивних 
розпадів за одну секунду.

При такому великому μ:
- Закон великих чисел працює виключно добре навіть для малих n
- Відносна дисперсія дуже мала: SD[S]/E[S] ≈ 3.5×10⁻⁵
- Сума S сильно сконцентрована навколо свого математичного сподівання
- Навіть найгрубіша нерівність Маркова дає досить точну оцінку

Тому всі три методи (включаючи грубу оцінку Маркова) збігаються до 
близьких значень n.

2. ЧОМУ ЕМПІРИЧНА ЙМОВІРНІСТЬ ДОРІВНЮЄ 1.0000?

Для n = 50:
- Очікувана сума: E[S] = 50 × μ ≈ 798,961,200
- Критичне значення: c = 800,000,000
- Різниця: c - E[S] ≈ 1,038,800 (позитивна!)
- Стандартне відхилення: SD[S] ≈ 28,266

Відстань від E[S] до критичного значення становить приблизно 36.75 
стандартних відхилень! Для нормального розподілу це означає, що 
ймовірність P(S > c) менша ніж 10⁻²⁹⁵. Тому у всіх 10,000 симуляцій 
сума виявилася меншою за критичне значення, і емпірична ймовірність 
склала 1.0000.

3. ЧОМУ ГРАФІК РІЗКО ПАДАЄ МІЖ n=50 ТА n=51?

Ключова різниця між n=50 та n=51:

При n = 50:
- E[S] ≈ 798,961,200 < 800,000,000 (менше за поріг)
- Сума S майже завжди буде менша за критичне значення

При n = 51:
- E[S] ≈ 814,940,424 > 800,000,000 (більше за поріг!)
- Сума S майже завжди буде більша за критичне значення

Різниця в одиницю n змінює очікувану суму на μ ≈ 16 мільйонів, що 
переміщує розподіл з одного боку порогу на інший. Через велике μ 
та малу відносну дисперсію розподіл S дуже вузький, тому перехід 
відбувається практично миттєво.

4. ВЕРИФІКАЦІЯ СИМУЛЯЦІЄЮ:

Симуляція з K = 10,000 повторень підтвердила теоретичні результати:
- Для n = 50: емпірична P(S ≤ c) = 1.0000 (умова виконана)
- Для n = 51: емпірична P(S ≤ c) ≈ 0 (умова не виконана)

Це повністю узгоджується з теоретичними розрахунками за всіма трьома 
методами.

ВИСНОВОК:
---------
Максимальне значення n = 50 зразків.

Близькість результатів усіх трьох методів (Марков, Чернов, ЦГТ) 
пояснюється виключно великим значенням параметра μ ≈ 1.6×10⁷, що 
забезпечує сильну концентрацію суми навколо математичного сподівання 
та робить навіть грубі оцінки достатньо точними. Симуляційна перевірка 
повністю підтвердила теоретичні результати.
============================================================

**Next, proceed with all the remaining steps**

**Do not forget to include several sentences summarizing your work and the conclusions you have made!**

------------------------------------------------------------------------

### Task 3.

#### In this task, we use the Central Limit Theorem approximation for continuous random variables.

#### One of the devices to measure radioactivity level at a given location is the Geiger counter. When the radioactive level is almost constant, the time between two consecutive clicks of the Geiger counter is an exponentially distributed random variable with parameter $\nu_1 = \mathtt{team\,id\,number} + 10$. Denote by $X_k$ the random time between the $(k-1)^{\mathrm{st}}$ and $k^{\mathrm{th}}$ click of the counter.

1.  Show that the distribution of the sample means of $X_1, X_2,\dots,X_n$ gets very close to a normal one (which one?) as $n$ becomes large. To this end,
    -   simulate the realizations $x_1,x_2,\dots,x_n$ of the \textbf{r.v.} $X_i$ and calculate the sample mean $s=\overline{\mathbf{x}}$;\
    -   repeat this $K$ times to get the sample $\mathbf{s}=(s_1,\dots,s_K)$ of means and then the \emph{empirical cumulative distribution} function $F_{\mathbf{s}}$ of $\mathbf{s}$;\
    -   identify $\mu$ and $\sigma^2$ such that the \textbf{c.d.f.} of $\mathscr{N}(\mu,\sigma^2)$ is close to the \textbf{e.c.d.f.} $F_{\mathbf{s}}$ of and plot both \textbf{c.d.f.}'s on one graph to visualize their proximity;\
    -   calculate the maximal difference between the two \textbf{c.d.f.}'s;\
    -   consider cases $n = 5$, $n = 10$, $n=50$ and comment on the results.
2.  The place can be considered safe when the number of clicks in one minute does not exceed $100$. It is known that the parameter $\nu$ of the resulting exponential distribution is proportional to the number $N$ of the radioactive samples, i.e., $\nu = \nu_1*N$, where $\nu_1$ is the parameter for one sample. Determine the maximal number of radioactive samples that can be stored in that place so that, with probability $0.95$, the place is identified as safe. To do this,
    -   express the event of interest in terms of the \textbf{r.v.} $S:= X_1 + \cdots + X_{100}$;\
    -   obtain the theoretical bounds on $N$ using the Markov inequality, Chernoff bound and Central Limit Theorem and compare the results;\
    -   with the predicted $N$ and thus $\nu$, simulate the realization $x_1,x_2,\dots,x_{100}$ of the $X_i$ and of the sum $S = X_1 + \cdots + X_{100}$;\
    -   repeat this $K$ times to get the sample $\mathbf{s}=(s_1,\dots,s_K)$ of total times until the $100^{\mathrm{th}}$ click;\
    -   estimate the probability that the location is identified as safe and compare to the desired level $0.95$

#### First, generate samples an sample means:

```{r}
nu1 <- 1  # change this!
K <- 1e3
n <- 5
sample_means <- colMeans(matrix(rexp(n*K, rate = nu1), nrow=n))
```

#### Next, calculate the parameters of the standard normal approximation

```{r}
mu <- 0       # change this!
sigma <- 1    # change this!
```

#### We can now plot ecdf and cdf

```{r}
xlims <- c(mu-3*sigma,mu+3*sigma)
Fs <- ecdf(sample_means)
plot(Fs, 
     xlim = xlims, 
     col = "blue",
     lwd = 2,
     main = "Comparison of ecdf and cdf")
curve(pnorm(x, mean = mu, sd = sigma), col = "red", lwd = 2, add = TRUE)
```

**Next, proceed with all the remaining steps**

**Do not forget to include several sentences summarizing your work and the conclusions you have made!**

------------------------------------------------------------------------

### Task 4.

**This task consists of two parts:**

1.  **In this part, we discuss independence of random variables and its moments: expectation and variance.**

    1.  Suppose we have a random variable $X$. Explain why $\mathbb{E}(\frac{1}{X}) \neq \frac{1}{\mathbb{E(X)}}$;

    2.  Let $X \sim \mathscr{N}(\mu,\sigma^2)$ with $\mu = teamidnumber$ and $\sigma^{2} = 2\times teamidnumber+7$. Simulate realizations $x_1,x_2,\dots,x_{100}$ of $X$ and $y_1,y_2,\dots,y_{100}$ of $Y := \frac{1}{X}$ to calculate the values of $\frac{1}{\overline{\textbf{X}}}$ and $\overline{\textbf{Y}}$. Comment on the received results;

        ```{r}
        mu <- 0 #change this!
        sigma <- 1 #change this
        N <- 100

        # YOUR CODE HERE
        ```

    3.  Let $X$ and $Y$ be exponentially distributed r.v.'s with parameter $\lambda = 2$. Set $Z := \log{X} + 5$. Plot the Quantile-Quantile plot and scatterplot of $X$ and $Y$. Plot the Quantile-Quantile plot and scatterplot of $X$ and $Z$. Explain the results. Comment on the difference of relations between the pairs of random variables. Which pair of r.v.'s is dependent and which one is similar?

        ```{r}
        # YOUR CODE HERE
        ```

    ------------------------------------------------------------------------

2.  You toss a fair coin three times and a random variable $X$ records how many times the coin shows Heads. You convince your friend that they should play a game with the following payoff: every round (equivalent to three coin tosses) will cost £$1$. They will receive £$0.5$ for every coin showing Heads. What is the expected value and the variance of the random variable $Y := 0.5X-1$?

    To answer this,

    1.  Explain what type of random variable is X:

        -   Normally distributed

        -   Binomially distributed

        -   Poisson distributed

        -   Uniformly distributed

    2.  What are the expected value and variance of X? Simulate realizations $x_1,x_2,\dots,x_{100}$ of $X$ to calculate the values of sample mean $\overline{\mathbf{X}}$ and sample variance $s^2 = \frac{\sum_{i=1}^{n}{(x_i - \overline{x})^{2}}}{n-1}$. Comment on the results;

        ```{r}
        # YOUR CODE HERE
        ```

    3.  What are the expected value and variance of Y? Simulate realizations $y_1,y_2,\dots,y_{100}$ of $Y$ to calculate the values of sample mean $\overline{\mathbf{Y}}$ and sample variance $s^2 = \frac{\sum_{i=1}^{n}{(y_i - \overline{y})^{2}}}{n-1}$. Comment on the results;

        ```{r}
        # YOUR CODE HERE
        ```

**Do not forget to include several sentences summarizing your work and the conclusions you have made!**

------------------------------------------------------------------------

### General summary and conclusions

Summarize here what you've done, whether you solved the tasks, what difficulties you had etc.

